Statistical framework for whole-genome bisulfite-sequencing data

### 2014-12-23 Rmarkdown/comethylation.Rmd
Minor fixes

In this chapter I describe two measures of within-sample co-methylation, _within-fragment co-methylation_ and _correlation of $\beta$-values_. Using both these measures I explore how co-methylation is affected by the distance between methylation loci, the density of methylation loci and the genomic context of the methylation loci. I apply these methods to 40 whole-genome bisulfite-sequencing samples to study the between-sample variability of co-methylation.

Throughout this chapter I focus on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci. The reason I do not explore co-methylation at non-CpG methylation loci is because such loci are rarely methylated in the samples that I am studying.

@@

Chapter \ref{chap:co-methylation_review} documented several previous analyses based on 'correlations' of aggregate methylation levels, such as $\beta$-values. What was lacking from these were clear descriptions of the proposed methods and software implementations.

@@

The first strategy creates all pairs of adjacent CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context. For a chromosome with $N$ CpGs there are $(N - 1)$ pairs. I call this the 'adjacent pairs' or $NIL = 0$ strategy, since all pairs have zero intervening loci^[nil0].

[^nil]: Recall that this is with respect to the reference genome and not the sample genome.

The second strategy uses all pairs of CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context. In practice, the second strategy only uses pairs for a set of $IPD$s, e.g., $IPD = 2 - 2000$, otherwise the number of pairs becomes unwieldy. This is more like a traditional autocorrelation analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$, albeit of a vector of unequally spaced observations and with care taken to only use pairs on the same chromosome and strand. I refer to this as the $NIL \geq 0$ strategy. Using the $NIL \geq 0$ strategy, two pairs of CpGs with an identical $IPD$ may have a very different number of intervening CpGs.

The interpretation of correlations computed using the $NIL \geq 0$ is complicated because the set of pairs are more heterogeneous than those under the $NIL = 0$ strategy. For this reason I prefer the $NIL = 0$ strategy, however, it may be useful to consider and contrast the two strategies, which is what I have done in my thesis.

@@

Even when using the $NIL = 0$ strategy, two pairs of CpGs with an identical $IPD$ may come from two distinct regions of the genome that have very different methylation dynamics. In other words, the vector of $\beta$-values for a given sample is highly non-stationary --- the covariance and correlation between two loci depends on more than just the $IPD$. An obvious thing to investigate is how strong an influence the genomic context has on these correlations of $\beta$-values. To do this, we can stratify our analysis by a genomic feature. While this does not eliminate this non-stationarity, it may help identify factors that drive some of the heterogeneity.

@@

The three commonly used correlation coefficients are Pearson's $r$ (__CITE__), Spearman's $\rho$ (__CITE__) and Kendall's $\tau$ (__CITE__). Briefly, Pearson's product-moment correlation is designed to detect linear relationships between two variables. However, it is not robust to outliers, which motivates the use of Spearman's and Kendall's rank-based correlation coefficients. Spearman's $\rho$ is simply the Pearson product-moment correlation of the ranks of the data while Kendall's $\tau$ is based on the relative frequency of concordantly ranked pairs.

@@

To address this problem, we might try smoothing the raw correlations, for example, by using LOESS (__CITE__), and then looking for periodicities in the smoothed values. However, this is only helpful when the data are corrupted by a small amount of noise, certainly not the extreme noise we see for $IPD > 1000$ in most samples. Furthermore, smoothing introduces problems of its own, such as choosing the smoothing bandwidth in such a way as to not introduce or destroy real periodicities. Even if we can automatically choose these smoothing parameters in an unbiased manner, we are still stuck with only observing a handful of periods of this size from $NIL = 0$ pairs. So while these periodicities are 'obvious' in the plots, they are rather more difficult to detect without visualising the raw data.

@@

Firstly, the ``possible range of the [log] odds ratio is unaffected by the marginal distributions'' (__CITE: quote from \cite{Kraemer:2004el} but references two other papers. Find original quote__). Another way of putting this is that given an odds ratio and a set of marginal probabilities, there exists one and only one $2 \times 2$ table satisfying these constraints (__CITE: Plackett?__). In contrast, Pearson's correlation of a $2 \times 2$ is restricted by the marginal probabilities (__CITE: Kraemer HC. A measure of 2 × 2 association with stable variance and approximately normal small sample distribution: planning cost-effective studies.__), which I find undesirable.

>>>>

In this chapter I describe two measures of within-sample co-methylation, _within-fragment co-methylation_ and _correlation of $\beta$-values_. Using both these measures I explore how co-methylation is affected by the distance between methylation loci and the genomic context of the methylation loci. I apply these methods to 40 whole-genome bisulfite-sequencing samples to study the between-sample variability of co-methylation.

Throughout this chapter I focus on co-methylation of CpGs, although the methods described are also applicable to non-CpG methylation loci. I do not explore co-methylation of non-CpG methylation loci because these loci are rarely methylated in the samples that I am studying.

@@

Chapter \ref{chap:co-methylation_review} documented several previous analyses based on 'correlations' of aggregate methylation levels, such as $\beta$-values. What was lacking from these were clear descriptions of the proposed methods and software implementations. In this chapter I try to rectify the former and with the `MethylationTuples` I make available a software implementation of these methods.

@@

The first strategy creates all pairs of adjacent CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context. For a chromosome with $N$ CpGs there are $(N - 1)$ pairs. I call this the 'adjacent pairs' or $NIL = 0$ strategy, since all pairs have no intervening loci ($NIL$ = number of intervening loci)^[nil0].

[^nil0]: Recall that this is with respect to the reference genome and not the sample genome.

The second strategy uses all pairs of CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context. In practice, the second strategy only uses pairs from a set of $IPD$s, e.g., $IPD = 2 - 2000$, otherwise the number of pairs becomes unwieldy. This is more like a traditional autocorrelation analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$, albeit of a vector of unequally spaced observations and with care taken to only use pairs on the same chromosome and strand. I refer to this as the $NIL \geq 0$ strategy. Using the $NIL \geq 0$ strategy, two pairs of CpGs with an identical $IPD$ may have a very different number of intervening CpGs.

The interpretation of correlations computed using the $NIL \geq 0$ is complicated because the set of pairs are more heterogeneous than those under the $NIL = 0$ strategy. For this reason I prefer the $NIL = 0$ strategy, however, it is useful to consider and contrast the two strategies, which is what I have done in my thesis.

@@

Even when using the $NIL = 0$ strategy, two pairs of CpGs with an identical $IPD$ may come from two regions of the genome that have very different methylation dynamics. In other words, the vector of $\beta$-values for a given sample is highly non-stationary --- the correlation between two loci depends on more than just the $IPD$. An obvious thing to investigate is how strong an influence the genomic context has on these correlations of $\beta$-values. To do this, we can stratify our analysis by a genomic feature. While this does not eliminate this non-stationarity, it may help identify factors that drive some of the heterogeneity.

@@

Three commonly used correlation coefficients are Pearson's $r$ \citep{Pearson:1895vt}, Spearman's $\rho$ \citep{Spearman:1904eg} and Kendall's $\tau$ \citep{KENDALL:1938bv}. Briefly, Pearson's product-moment correlation is designed to detect linear relationships between two variables. However, it is not robust to outliers, which motivates the use of Spearman's and Kendall's rank-based correlation coefficients. Spearman's $\rho$ is simply the Pearson product-moment correlation of the ranks of the data while Kendall's $\tau$ is based on the relative frequency of concordantly ranked pairs.

@@

To address this problem, we might try smoothing the raw correlations, for example, by using LOESS \citep{Cleveland:1979fu}, and then looking for periodicities in the smoothed values. However, this is only helpful when the data are corrupted by a small amount of noise, certainly not the extreme noise we see for $IPD > 1000$ in most samples. Furthermore, smoothing introduces problems of its own, such as choosing the smoothing bandwidth in such a way as to not introduce or destroy real periodicities. Even if we can automatically choose these smoothing parameters in an unbiased manner, we are still stuck with only observing a handful of periods of this size from $NIL = 0$ pairs. So while these periodicities are 'obvious' in the plots, they are rather more difficult to detect without visualising the raw data.

@@

Firstly, the ``possible range of the [log] odds ratio is unaffected by the marginal distributions'' \citep{Kraemer:2004el}. Another way of putting this is that given an odds ratio and a set of marginal probabilities, there exists one and only one $2 \times 2$ table satisfying these constraints \citep{plackett1981analysis}. In contrast, Pearson's correlation of a $2 \times 2$ is restricted by the marginal probabilities \citep{kraemer1986measure}, which I find undesirable.

### 2014-12-23 Rmarkdown/wgbs_downstream_analyses.Rmd
Latex fixes
Minor fixes

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
& Allele 1 & Allele 2 \\ \midrule
m &  m_{1}        &  m_{2}       \\
u &  u_{1}        &  u_{2}       \\ \bottomrule
\end{tabular}
\caption{$2 \times 2$ table used to test for allele-specific methylation.}
\label{tab:asm}
\end{table}

@@

Regardless of where you draw the line as to what constituted an epiallele, it has become clear that the occurrence of multiple methylation patterns at an m-tuple is the norm rather than the exception when performing bisulfite-sequencing. For example, __FIGURE__ shows the number of methylation patterns observed at 4-tuples with a frequency greater than __WHAT__ from the __WHICH__ sample. There has been some progress in developing software to extract and analyse these epialleles, which I now describe. (Alternatively, cite Landan et al.)

>>>>

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
& Allele 1 & Allele 2 \\ \midrule
$m$ &  $m_{1}$        &  $m_{2}$       \\
$u$ &  $u_{1}$        &  $u_{2}$       \\ \bottomrule
\end{tabular}
\caption{$2 \times 2$ table used to test for allele-specific methylation.}
\label{tab:asm}
\end{table}

@@

Regardless of where you draw the line as to what constituted an epiallele, it has become clear that the occurrence of multiple methylation patterns at an m-tuple is the norm rather than the exception when performing bisulfite-sequencing \citep{Landan:2012kp}. There has been some progress in developing software to extract and analyse these epialleles, which I now describe.

### 2014-12-23 Rmarkdown/wgbs_statistical_framework.Rmd
Added section label
Figure fixing.
Minor fixes

Finally, I describe key variables, common estimators of these and their statistical properties by analysing __HOW MANY__ whole-genome bisulfite-sequencing samples.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/bs-seq_protocol.pdf}
\caption{A schematic illustrating several sources of variation in a bisulfite-sequencing experiment. Each sample starts as a population of cells, with potentially different methylation patterns. Each DNA fragment is coloured by its originating cell, although this is unknown in practice. Illustrated are PCR amplication bias (unequal representation of DNA from each cell following PCR amplification), sampling variation (not all DNA fragments are sequenced), alignment errors (not all sequenced fragments are mapped or may be mapped to the wrong location, as is the case for the green fragment), and filtering during methylation calling (not reads are used, the blue read, and reads may have read-positions removed, the black reads).}
\label{fig:bs-seq_protocol}
\end{figure}

@@

\subsubsection{Genome-wide distribution of $\beta$-values}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:EPISCOPE_plotBetaDensity}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Lister_plotBetaDensity}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Seisenberger_plotBetaDensity}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Ziller_merged_plotBetaDensity}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:EPISCOPE_plotBetaFreqPoly}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:Lister_plotBetaFreqPoly}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:Seisenberger_plotBetaFreqPoly}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Ziller data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:Ziller_merged_plotBetaFreqPoly}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:EPISCOPE_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Lister_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Seisenberger_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Ziller_merged_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
\label{fig:EPISCOPE_plotBetaFreqPoly_CGI}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
\label{fig:Lister_plotBetaFreqPoly_CGI}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
\label{fig:Seisenberger_plotBetaFreqPoly_CGI}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Zilelr data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
\label{fig:Ziller_merged_plotBetaFreqPoly_CGI}
\end{figure}

>>>>

Finally, I describe key variables, common estimators of these and their statistical properties by analysing 40 whole-genome bisulfite-sequencing samples.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/bs-seq_protocol.pdf}
\caption{A schematic illustrating several sources of variation in a bisulfite-sequencing experiment. Each sample starts as a population of cells, with potentially different methylation patterns. Each DNA fragment is coloured by its originating cell, although this is unknown in practice. Illustrated are PCR amplication bias (unequal representation of DNA from each cell following PCR amplification), sampling variation (not all DNA fragments are sequenced), alignment errors (not all sequenced fragments are mapped or may be mapped to the wrong location, as is the case for the green fragment), and filtering during methylation calling (not reads are used, the blue read, and reads may have read-positions removed, the black reads).}
\label{fig:bs-seq_protocol}
\end{figure}

@@

\subsubsection{Genome-wide distribution of $\beta$-values}\label{sec:genome-wide_distribution_of_beta_values}

@@

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:EPISCOPE_plotBetaDensity}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Lister_plotBetaDensity}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Seisenberger_plotBetaDensity}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data. Densities were normalised so that the maximum value for each sample is 1. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. `Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Ziller_merged_plotBetaDensity}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:EPISCOPE_plotBetaFreqPoly}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:Lister_plotBetaFreqPoly}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:Seisenberger_plotBetaFreqPoly}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Ziller data. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included.}
\label{fig:Ziller_merged_plotBetaFreqPoly}
\end{figure}

@@

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:EPISCOPE_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Lister data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Lister_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Seisenberger_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaDensity_CGI.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the Ziller data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Spikes' in the density estimate are due to the discreteness of $\beta$-values.}
\label{fig:Ziller_merged_plotBetaDensity_CGI}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
\label{fig:EPISCOPE_plotBetaFreqPoly_CGI}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Lister_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Lister data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
\label{fig:Lister_plotBetaFreqPoly_CGI}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Seisenberger_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Seisenberger data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
\label{fig:Seisenberger_plotBetaFreqPoly_CGI}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{../figures/Ziller_merged_plotBetaFreqPoly_CGI.pdf}
\caption{Frequency polygon of the genome-wide distribution of CpG $\beta$-values for the Zilelr data, stratified by whether the CpG is in a CpG island. $\beta$-values are grouped into $0.01$-width bins and the percentage of CpGs in each bin is plotted on the y-axis. Only CpGs with at least $10\times$ sequencing coverage are included. Percentages are with respect to all CpGs with at least $10 \times$ sequencing coverage, unstratified by CpG island status.}
\label{fig:Ziller_merged_plotBetaFreqPoly_CGI}
\end{figure}
